---
title: "forcats::fct_rev()"
subtitle: "Function of the Week"
description: "fct_rev(): Reverse factor levels"
author: "Michael Daily"
date: "3/5/2025"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 4
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor_options: 
  chunk_output_type: inline
---
 

# Factor Reverse: fct_rev()

This document discusses fct_rev() and its use. fct_rev() is from the forcats package and, similar to other functions from forcats, is a tool for handling factor type variables. 

```{r}
#load packages
library(tidyverse)
library(datasets)
library(forcats)
library(janitor)
#ChickWeight dataset
chick_wgt <-ChickWeight %>% clean_names()
glimpse(chick_wgt)
```

## What does it do and what is it for?

### What fct_rev() does:

fct_rev() does what it sounds like. It reverses the levels of the factor variable. It takes only one argument which is a factor (or character vector). 

```{r}
#look at the levels of diet
levels(chick_wgt$diet)
#revers the levels
levels(fct_rev(chick_wgt$diet))
```


### What fct_rev() is useful for: 

fct_rev() is useful for plotting factors on the y-axis. By default, the levels are plotted with the first level at the bottom and the last at the top as shown below.  

```{r}
ggplot(chick_wgt) + 
  aes(y=diet,
      x=weight,
      fill=diet) +
  geom_boxplot() +
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.y=element_blank(),
        plot.title=element_text(hjust=.5))+
  labs(y='Diet', x='Weight', fill='Diet', title='Weight Distribution per Diet (original order)')
  
```

fct_rev() makes it easy to plot the reverse with the first level at the top and the last level at the bottom as shown in the following figure. 

```{r}
ggplot(chick_wgt) + 
  aes(y=fct_rev(diet),
      x=weight,
      fill=diet) +
  geom_boxplot() +
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.y=element_blank(),
        plot.title=element_text(hjust=.5))+
  labs(y='Diet', x='Weight', fill='Diet', title='Weight Distribution per Diet (reversed order)')
```




## Is it helpful?

In cases where you want to plot a factor on the y-axis, this is a helpful function; this is particularly true when the factor variable has many levels and it would be tedious to reorder by hand. Aside from reordering levels for a figure, this function seems like it would rarely, if ever, be useful. 
