---
title: "stringr::str_trim"
subtitle: "Function of the Week"
description: "Removes whitespace from the begining and/or end of a string"
author: "Destine Krenik"
date: "2/27/25"
format: 
  html:
    theme: superhero
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 4
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor_options: 
  chunk_output_type: inline
---


```{r, include=FALSE}

#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
```


# str_trim()

In this document, I will introduce the str_trim function and show what it's for.

```{r}
#load libraries
library(tidyverse)
library(stringr)
```

## What is it for?

The str_trim() function removes whitespace from the begining and/or end of a string. You can specify which end by setting "side" equal to "left", "right", or "both".

```{r}
#Create example data frame
id <- c(1:3)
name_animal <- c(" Jeff bear", "Anne penguin   ", "       Richard lion ")

df1 <- data.frame(id, name_animal)
df2 <- data.frame(id, name_animal)
df3 <- data.frame(id, name_animal)

print(df1)
```
```{r}
#What happens if you try to separate the column name_animal?
df_sep <- df1 %>%
  separate(col = name_animal,
           into = c("name", "animal"),
           sep = " ")

print(df_sep)
```

```{r}
#Trim whitespace from the beginning
str_trim(df1$name_animal, side = "left")

#Trim whitespace from the end
str_trim(df2$name_animal, side = "right")

#Trim whitespace from both sides
str_trim(df3$name_animal, side = "both")
```
```{r}
#Now what happens when we try to separate our cleaned vector?
df3 <- df3 %>%
  mutate(name_animal = str_trim(name_animal))

df_sep <- df3 %>%
  separate(col = name_animal,
           into = c("name", "animal"),
           sep = " ")

print(df_sep)
```


## Is it helpful?

This function can be very helpful when cleaning data. However, there is another function that may be better for general use called str_squish(). The str_squish() function removes all excess whitespace from the beginning and end of a string, and replaces all whitespace between words with a single space. 
