---
title: "dplyr::between()"
subtitle: "Function of the Week"
description: "Detect where values fall in a specified range"
author: "Sophia Mann"
date: "2/11/2025"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 4
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor_options: 
  chunk_output_type: inline
---

```{r, include=FALSE}

#| label: setup
#| include: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

# between( )

In this document, I will introduce the between( ) function and show what it can be used for.

```{r}

library(tidyverse)

library(palmerpenguins)
data(penguins)
```

## What is it for?

The between( ) function from the dplyr package makes it easier to find values within a specified range (inclusive of the lower and upper bounds).

### Example 1 

```{r}

#1  getting info on penguin body mass 
summary(penguins$body_mass_g)

#2  Getting values that are between the 1st and 3rd quartiles using >= , <= 

penguins_filtered <- penguins %>%
  filter(body_mass_g>= 3550  & body_mass_g <= 4750)

head(penguins_filtered)

# Getting values that are between the 1st and 3rd quartiles using between()

penguins_filtered_btwn<- penguins %>%
  filter(between(body_mass_g, 3550, 4750))

head(penguins_filtered_btwn)


```

### Example 2 

```{r}
#1  getting info on penguin bill length  
summary(penguins$bill_length_mm)

#2 creating dataset of penguins with shirt beaks 

Penguins_short_beaks <- penguins%>% 
  filter(between(bill_length_mm, 32.10, 39.23))

head(Penguins_short_beaks)
  
```

## Is it helpful?

> **Discuss whether you think this function is useful for you and your work. Is it the best thing since sliced bread, or is it not really relevant to your work?**

Yes, it works well with other functions such as mutate or filter, and also makes the readability of your code better. Readability of your code is important because it makes it easier for you to read and find errors, and makes it easier for others to work with your code.
