---
title: "ggplot2::geom_tile"
subtitle: "Function of the Week"
description: "Used for data visualization, specifically in the form of heatmaps or other tile plots"
author: "Sanya Surya"
date: "1/30/25"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 4
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}

#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
```

# geom_tile()

In this document, I will introduce the geom_tile() function and show what it's for.

```{r}
# Loading tidyverse and other libraries
pacman::p_load(
  tidyverse,    
  readxl,       
  ggthemes,     
  ggplot2,
  here
  )
```

## What is it for?

-   This function comes from the ggplot2 library and is used for data visualization, specifically in the form of heatmaps or other tile plots.

-   Heatmaps are a useful visual way to show the intensity of a numerical variable in the context of two categorical variables.

-   In the code below, I will show how geom_tile() can be used to create a heatmap looking at a data set of HIV Diagnoses and AIDS Deaths in Oregon from 2012 to 2022. We will be looking at the count of cases of these indicators.

-   *Data Source: Retrieved from CDC.gov.*

-   *To get a CSV file, which I converted to an Excel file, I chose STI data in the form of a table and specified parameters: Indicators: AIDS Deaths, HIV Diagnoses, Age: 13+, Sex: Both, State: Oregon, Years: 2012-2022, All transmission categories, All races and ethnicities.*

```{r}

# Load and view the STI Dataset
# Skip the first 11 lines to reach the headers
OR_STI <- read_excel(here("function_week", "data", "data_SURYA.xlsx"), sheet = 1, skip = 11, na = "NA")
OR_STI

```

```         
```

```{r}
# Create the heatmap
myplot <- ggplot(OR_STI,
       aes(x = Year,
           y = Indicator,
           fill = Cases)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  # Customize the angle and positions of x and y tick mark labels for better visualization (especially because the 2020 label has a much longer name)
  theme(axis.text.y = element_text(angle = 90, size = 8, hjust = 0.5),
        axis.text.x = element_text(angle = 25, hjust = 1, vjust = 1, size = 8)) +
  # Add a graph title
  labs(title = "Count of Cases of HIV and AIDS Indicators in Oregon from 2012 to 2022")
myplot
```

## Is it helpful?

-   Heatmaps are helpful to visualize categorical variables with a third numeric variable

-   When there is a greater range of values, such as in this dataset, heatmaps can show intensity in an easy-to-digest visual way as opposed to other options where readers would need to read numeric scales on a graph.

-   Can be helpful to visually stack and compare variables. In this data for instance, we can visually see that there have historically been a lot more HIV diagnoses than AIDS deaths, and HIV diagnoses fluctuate often with time. However, in recent years we can see a clear upward trend in counts of AIDS deaths since the color becomes lighter, indicating increasing cases.
